Emulating Comparative Oncology Trials with Real-world Evidence Studies (ENCORE): Process Development and Methodological Considerations for Oncology Real-World Data

### 2025-07-06 manuscript/manuscript.qmd
added FDA round 2 comments and edits

---
title: "A Process for Emulating Comparative Oncology Trials with Real-world Evidence Studies (ENCORE): Process Development and Methodological Considerations for Oncology Real-World Data"

@@

**Authors**: Janick Weberpals^1^, Sebastian Schneeweiss^1^, Kenneth L. Kehl^2^, Donna R. Rivera^3^, Pallavi Mishra-Kalyani^3^, Catherine C. Lerro^3^, Erin Larkins^4^, Sonia Singh^4^, Preeti Narayan^4^, Richard Curley^4^, Paul Kluetz^4^, Georg Hahn^1^, Priyanka Anand^1^, Yanina Natanzon^5^, Andrew J. Belli^6^, Ching-Kun Wang^6^, Jenna Collins^7^**,** Jonathan Kish^7^**,** Janet Espirito^8^, Nicholas J Robert^8^, Robert J. Glynn^1^, Shirley V. Wang^1^

@@

@@

Real-world evidence (RWE) is increasingly used to complement findings from randomized controlled trials (RCTs), contextualizing the effectiveness and safety of medical interventions as delivered in routine clinical practice. Advancements in the curation and accessibility of electronic health record data (EHR) present the opportunity to utilize real-world data (RWD) to investigate therapeutic areas including oncology, where administrative healthcare claims databases alone are often not fit-for-purpose. The RCT DUPLICATE initiative has previously enhanced understanding of when RWE can most appropriately draw causal conclusions by emulating trials in non-oncology indications. Here, we present the design and trial selection for the Emulation of Comparative Oncology Trials with Real-world Evidence (ENCORE) project, which extends this work to oncology. ENCORE is designed to emulate 12 RCTs in four oncology-specialized EHR databases across four different cancer indications, specifically non-small cell lung cancer, breast cancer, colorectal cancer, and multiple myeloma. It will place a special emphasis on systematic evaluation of fitness of data in relation to the study design and statistical analysis for a particular research question, and pre-registration of study protocols prior to initiation and analysis. Agreement of treatment effect estimates between RCTs and their respective emulations will be assessed using pre-specified criteria. Through extensive sensitivity analyses benchmarked against RCT results, the ENCORE project will inform understanding of how measurement, design, and analytic decisions influence the validity of oncological RWE studies.

@@

Randomized controlled trials (RCTs) are the go-to methodology for establishing the efficacy and safety of medical products. Directed by the 21^st^ Century Cures Act,[@RWEFDA] the Food and Drug Administration (FDA) established a framework to factor consideration of real-world evidence (RWE) generated from routine-care health data such as electronic health records (EHR) to evaluate and contextualize the safety and effectiveness of novel cancer therapies.[@purpura2022role] Accounting for 30% of new drug approvals, oncology was the disease area with the most FDA drug approvals in 2024[@senior2024fresh] and has many areas of high unmet medical need to treat serious conditions. Therefore, RWE has particularly important potential to complement evidence from RCTs in the field of oncology. Potential use cases include the assessment of effectiveness in specific patient populations that are underrepresented in RCTs, the construction of external control arms in single-arm trials that have feasibility or equipoise challenges, or the precision oncology-focused discovery of biomarkers among pan-tumor populations that harbor specific genomic and immuno-pathological signatures.

However, the validity and transportability of results derived between RWE studies and RCTs depends on multiple factors. Frequently referenced limitations include lack of baseline randomization, missing data, unmeasured confounding, small study sizes, data discontinuity,[@Merola2022; @joshua2022longitudinal] adoption of changes in guidelines in real-world care and the inability to measure and emulate common eligibility criteria, including prognostic factors, and standardized response assessments in real-world data (RWD).[@rider2024emulations] While examples of oncology trial emulations have been published,[@rider2024emulations; @merola2023aetion; @merola2024calibrating] a systematic and scaled approach to emulate a diverse set of oncology trials with multiple heterogeneous databases is necessary to gain confidence in the validity of RWE studies, evaluate regulatory considerations, and to provide context as to which questions can be validly answered.

The RCT DUPLICATE initiative[@wang2023emulation] increased our understanding of when RWE studies can come to causal conclusions on treatment effects by benchmarking results against RCTs under the assumption that each RCT finding reflects a causal treatment effect. In settings where the RCT designs could be emulated well, RWE studies were able to reach similar conclusions.[@heyard2024design] However, this prior work from RCT-DUPLICATE focused primarily on emulating trials in non-oncology settings using claims databases.

The *Emulation of Comparative Oncology Trials with Real-world Evidence* (ENCORE) project [@encoreFDA] aims to extend this work to oncology. Studies in oncology come with their own unique set of methodological challenges which must be systematically explored and understood. Building on a process co-developed with the FDA through RCT DUPLICATE,[@wang2023emulation] this expansion to oncology will emulate 12 randomized clinical trials using multiple specialty oncology EHR data sources. The process will emphasize transparency and include documented data fit-for-purpose assessment[@rivera2024oncology; @gatto2022structured] for each RWD source with respect to each trial emulation as well as extensive sensitivity analyses to assess robustness of findings.

The objectives of this project are to 1) develop state-of-the-art methodological approaches and 2) to apply them to create insights into the potential use of RWE to enhance regulatory science in oncology. To achieve these objectives, this demonstration project will systematically emulate 12 oncology trials across four cancer types and assess the agreement of treatment effect estimates between RCTs and their respective emulations.

Here, we describe the design and process for the selection of the 12 oncology RCTs, the assessment of the database quality and selection, protocol development and pre-registration, study design and statistical analysis, and pre-specified agreement metrics to evaluate the concordance between RCTs and emulations.

@@

The focus of ENCORE is to efficiently explore in what contexts RWE studies can or cannot yield similar results compared to RCTs. Therefore, the project emphasizes trials of therapies for common cancers for which there has been substantial therapeutic development in recent years. After review and collaboration with clinical and regulatory experts, four cancer indications were identified: non-small cell lung cancer, breast cancer, colorectal cancer and multiple myeloma. For each cancer type, we aim to conduct three trial emulations using multiple accessible databases (i.e., the total equaling 12 trial emulations x *n* databases which are found fit-for-purpose for each trial).

We used a systematic process for trial selection where the eligibility criteria are documented in CONSORT diagrams showing reasons for excluding RCTs for each cancer type. The search was conducted using the *Aggregated Analysis of ClinicalTrials.gov* (AACT) database which is a publicly available relational database developed and maintained by the Clinical Trials Transformation Initiative (CTTI) that contains information (protocol and result data elements) about studies registered on ClinicalTrials.gov.[@tasneem2012database] To identify eligible trials, we used a combined search query strategy of the National Library of Medicine (NLM)-controlled *Medical Subject Headings **(***MeSH) term and a free keyword search for the respective cancer indication in the *conditions*, *studies* and *detailed_descriptions* fields of each trial entry on ClinicalTrials.gov.

@@

-   Intervention model: parallel assignment

@@

-   Trial start in 2011 or later

-   Primary purpose was to study treatment efficacy

-   Overall survival must be one of the endpoints reported (either as hazard ratio or median overall survival time)

@@

The rationale and operationalization for each criterion is listed in detail in @tbl-criteria. We will mainly consider pivotal large late-phase interventional RCTs after 2011 because treatment guidelines among included cancer indications have undergone significant changes in recent years. Due to the rapid adoption of breakthrough therapies in routine care, few patients currently initiate outdated treatment regimens in current clinical practice. Conversely, we excluded trials with results published too recently to allow for enough data and follow-up time accrual in the real-world databases used for this project. We did not define a global cut-off as the requirements for follow-up time are different for each cancer type and population (e.g., advanced NSCLC versus early-stage breast cancer) and the decisions were made on a case-by-case basis.

Although there have been substantial methodological advancements to increase our understanding on the emulation and comparison of real-world progression-free survival (PFS) and objective response rates (ORR) to a RECISTv1.1[@eisenhauer2009new]-based PFS and ORR assessment in RCTs [@ThanCCR; @mckelvey2024evaluation], imaging-based evaluations still hold a level of granularity which is not well reflected in chart-abstracted assessments of a patient's progression in routine care.[@rwdRECIST; @rivera2022friends] The timing and cadence of intervals between progression assessments can differ between RCTs and routine care which may result in measurement error and bias.[@ackerman2024measurement] Given this, and the large number of other methodological challenges including missing data, small sample sizes, data discontinuity and rapidly changing guideline treatments, we focus on the emulation of overall survival (OS) as the endpoint of interest. Therefore, we include only trials that have reported overall survival (OS) as one of the pre-specified endpoints in the protocol.

While most trial eligibility criteria could be operationalized in an automated fashion, our final criterion was an assessment of emulation feasibility and clinical relevance. This criterion involved extensive human review. The fundamental points considered in this step include an initial feasibility assessment of the data fitness,[@rivera2022friends] involving an assessment of whether critical eligibility criteria (e.g., biomarker status) including prognostic factors (e.g., ECOG performance score) are measurable, and whether preliminary study size based on a rough estimation of the number of patients observed in the data with the combination of treatment regimen and line of treatment is reasonable for shortlisting the trial. Lastly, trial candidates were ranked and shortlisted into primary and runner-up candidates based on their clinical and regulatory relevance.

A list of tentative, shortlisted primary candidates is presented in @tbl-rcts. The corresponding selection process is illustrated in the CONSORT diagrams (**Supplementary Figures 1-4**) and runner up candidate trials are listed in **Supplementary Table 1**. Naturally, the majority of shortlisted trials cover advanced (locally advanced, inoperable, recurrent/progressive disease) or metastatic cancer populations because a large proportion of drug development efforts have recently focused on these settings. A key objective that we aim to explore with the selected trials is to achieve a better understanding how different disease settings (early, late), line settings (\[neo\]adjuvant, first line, advanced lines of therapy), therapy protocols (monotherapy, combination therapy) and clinical characteristics (all-comer versus complex genetic or immunological signatures) can be emulated using RWD. If more thorough feasibility assessments after applying all relevant eligibility criteria suggest that the threat of bias from mis-measurement of key study parameters or residual confounding remains high, or that the study size is not sufficient, then runner-up candidate trials will be considered instead.

@@

The ENCORE project will use data from four oncology-specific electronic health records (EHR)-derived RWD data sources (in alphabetical order): ConcertAI, COTA, Flatiron Health, and Ontada/McKesson. The included databases comprise the necessary information to study medical product effectiveness in oncology derived from US-based oncology EHRs. A detailed description and sampling methodology will be provided with each trial emulation protocol. For ENCORE, not all databases will be available for each cancer indication and the names of the databases will be blinded and referred to as ENCORE DataBase (EDB) 1, 2, 3 and 4 for the final reporting of results (the numbering does not coincide with the above order of mention of the databases). If more than one database is considered fit-for-purpose for a respective trial emulation, the best possible analytic model will be employed for each database separately and final treatment effect estimates will be pooled using a meta-analytic approach using fixed effects (weights reflect the inverse of the variance of each database study) and random effects (weights reflect the inverse of the variance of each database study plus an additional variance term that quantifies the assumed heterogeneity between databases) models.[@nikolakopoulou2014interpret]

@@

For each selected RCT, a detailed protocol, pre-specifying key elements of the trial emulation, will be developed using the HARPER protocol template[@wang2022harmonized] recommended for regulatory submissions of RWE studies[@guidelinegeneral] and will be registered on ClinicalTrials.gov after review by a clinical and FDA regulatory expert panel. Following the target trial emulation framework, we will provide an explicit statement and rationale on how each element will be emulated including database selection, covariate measurement, operationalization of key eligibility criteria, study design, data analysis and causal contrasts of interest.[@hernan2022target; @hernan2016specifying] Since it is common that oncology RCTs update OS estimates periodically based on accrued follow-up time, the protocol will give a brief summary of each emulated RCT and specify which target OS estimates will be used for agreement metric evaluation (see @sec-agreement-metrics). All eligibility criteria will be extracted based on publications, publicly available protocols, and statistical analysis plans of the selected RCT.

@@

One important aspect when emulating oncology trials is the choice and estimation of the appropriate estimand of interest.[@rufibach2018] In prior RCT-DUPLICATE efforts, a “while on treatment” strategy was chosen for database studies that were designed to emulate trials where the primary trial analysis was intention-to-treat (e.g., a “treatment policy” estimand). This estimand was chosen to mimic the high adherence typically observed in large cardiovascular clinical trials in the context of low adherence in clinical practice. However, a “while on treatment” estimand would be inappropriate when emulating oncology trials, not only because of highly informative censoring, but also because OS outcome analyses in oncology are typically “treatment policy” estimands that allow crossover.[@manitz2022estimands] We plan to focus on “treatment policy” estimands for the primary analyses in our trial emulations. However, allowing crossover can dilute treatment effects, moving point estimates toward the null. Therefore, differences in cross-over and discontinuation rates between the trial and emulation may still contribute to estimand differences. Recognizing these practical challenges, we will carefully characterize the patterns of discontinuation or cross-over in both the trials and the database studies designed to emulate them.

@@

Real-world data fitness and emulation feasibility for each shortlisted candidate trial will be assessed in multiple steps based on direction from of the oncology quality, characterization, and assessment of real-world data (Oncology QCARD) Initiative.[@rivera2022friends] The first step assesses if relevant variables like exposure, line of therapy, outcomes, and covariates are generally available, measured and operationalizable in routine-care. Since many oncological RCTs in recent years have focused on selected, biomarker-defined populations, nuances in measurement and operationalizability of specific biomarkers must be reflected to ensure a representative and large enough study population. For example, immune checkpoint inhibitors have significantly changed the cancer treatment landscape since the first approvals of the CTLA-4 inhibitor ipilimumab in 2011 and the PD-1 inhibitors pembrolizumab and nivolumab in 2014 in the United States. However, the operationalization of the expression of the PD-L1 biomarker in RCTs (e.g., as a percent staining, tumor proportion score or combined positive score) has evolved, and as such PD-L1 '*positivity'* may have different definitions across calendar years based on different cut-off values.

According to the *Structured Process to Identify Fit-For-Purpose Data* (SPIFD) framework[@gatto2022structured], the next step outlines how eligibility criteria will be ascertained using a color-coded heatmap that will indicate the level of confidence on how well each criterion can be emulated in each selected database. As there are general eligibility criteria in oncology clinical trials which either are infeasible to emulate (e.g., physician-assessed survival prognosis of x months) or that are clinically not relevant for the emulation of the trial (e.g., male patients should be willing to use barrier contraception), the study team will decide on key eligibility criteria for the emulation of the trial.

Additionally, we will provide conceptual and operationalized definitions on how exposures, outcomes and covariates are defined in each respective database. There will be special emphasis on how exposure, in context of disease and line of therapy settings, and the OS outcome are emulated. For all considered databases, OS is typically a composite measure that, depending on the underlying database, can be derived from different mortality sources (i.e., EHR documentation, social security death index, obituary and other linkages). Given that not all relevant sources that provide mortality data are synchronized and updated uniformly, sensitivity analyses with more conservative (i.e., earlier) censoring dates will be considered for each trial emulation to mitigate the potential impact of ghost-time bias.[@meyer2020open]

@@

Particularly when emulating pivotal trials of paradigm-changing treatments, multiple aspects need to be considered, such as the contemporaneity of the control cohort, the adoption rate of the novel medical product in routine care, the magnitude of the clinical treatment benefit and the rate in which patients discontinue or cross-over, which could dilute effect estimates.

To that end, comprehensive data explorations will be performed as part of the protocol development to contextualize these parameters and (if reported) draw comparisons to the emulated trial. Examples for such standard diagnostics are visualized in @fig-initiators. All exploratory analyses will be conducted blinded towards the outcome to avoid influencing study design and analytic choices.

The distribution of patient characteristics, by exposure status, will be examined using Table 1's before and after applying eligibility criteria and contrasted with the distributions of patient characteristics of the original RCT. Initial propensity score matching or weighting methods will be applied to ensure that measured pre-exposure covariates can be balanced, exposure cohorts are conditionally exchangeable at baseline and resulting sample sizes are still sufficient after matching or weighting. At this stage, all exploratory analyses will be conducted blinded towards the outcome to not bias any study design or analytic choices based on known outcome information.

@@

Causal analyses of observational data are often not designed with respect to formal hypothesis testing and statistical power in the same manner as RCTs since the number of 'recruited' patients is limited by available data and cannot be influenced.[@hernan2022causal] For this project, however, the emulations in EHR data must have at least equal power to the relevant trial for interpretation of the pre-specified agreements metrics between RCT and RWD results. Since the main outcome of interest is defined as time to all-cause mortality (OS), the estimation of the statistical power is driven by the number of events rather than the number of patients. To assess if the overall number of events, unstratified by exposure, is sufficient such that a significant difference can be detected based on the original RCT-reported hazard ratio (HR), the statistical power (1-$\beta$) will be estimated using Schoenfeld's sample-size formula for the proportional-hazards regression model.[@schoenfeld1981asymptotic]

@@

To formally compare treatment effects between RCTs and their respective emulations, we will adapt the approach of the RCT-DUPLICATE project.[@franklin2020nonrandomized; @wang2023emulation] That is, for the primary endpoint of interest (HR for OS and corresponding 95% confidence intervals), we will derive three qualitative agreement metrics: statistical agreement, estimate agreement and agreement based on the standardized mean difference (SMD). Examples are illustrated in @tbl-metrics.

**Statistical significance agreement**: agreement between RCT and emulated trial treatment effect with regards to directionality and statistical significance.

**Estimate agreement**: agreement that the estimated RWE treatment effect is within the 95% CI of the RCT treatment effect estimate. Provided that for some emulations, the power of the RWE study may be larger than that of the original RCT, this could lead to situations where there is no statistical significance agreement although the treatment effect estimates are highly overlapping but with the RCT estimate crossing the null (or vice versa in case the RCT has a larger power than the RWE emulation).

@@

The study design for each trial emulation will be visualized as part of the protocol using a graphical depiction of the exact measurement windows of eligibility criteria, washout periods and covariates relative to the cohort entry time.[@schneeweiss2019graphical]

@@

Due to its ubiquity in oncology trials, the primary parameter of interest in ENCORE will be defined as the marginal hazard ratio (HR) coefficient for the treatment comparison for OS.[@cox1972regression] We will also consider alternative endpoints on an absolute risk scale such as median survival times, survival probabilities at pre-defined time points during follow-up[@kaplan1958nonparametric] and restricted mean survival times as secondary endpoints of interest.

>>>>

---
title: "Emulating Comparative Oncology Trials with Real-world Evidence Studies (ENCORE): Process Development and Methodological Considerations for Oncology Real-World Data"

@@

**Authors**: Janick Weberpals^1^, Sebastian Schneeweiss^1^, Kenneth L. Kehl^2^, Donna R. Rivera^3^, Pallavi Mishra-Kalyani^3^, Catherine C. Lerro^3^, Erin Larkins^4^, Sonia Singh^4\*^, Preeti Narayan^4^, Richard Curley^4^, Georg Hahn^1^, Priyanka Anand^1^, Yanina Natanzon^5^, Andrew J. Belli^6^, Ching-Kun Wang^6^, Jenna Collins^7^**,** Jonathan Kish^7^**,** Janet Espirito^8^, Nicholas J Robert^8^, Robert J. Glynn^1^, Shirley V. Wang^1^

@@

\*At the time this research was conducted

@@

Real-world evidence (RWE) is increasingly used to complement findings from randomized controlled trials (RCTs), contextualizing the effectiveness and safety of medical interventions as delivered in routine clinical practice. Advancements in the curation and accessibility of electronic health record data (EHR) present the opportunity to utilize real-world data (RWD) to investigate therapeutic areas including oncology, where administrative healthcare claims databases alone are often not fit-for-purpose. The RCT DUPLICATE initiative has previously enhanced understanding of when RWE can most appropriately draw causal conclusions by emulating trials for non-oncology indications. Here, we present the design and trial selection for the Emulation of Comparative Oncology Trials with Real-world Evidence (ENCORE) project, which extends this work to oncology. ENCORE is designed to emulate 12 RCTs in four oncology-specialized EHR databases across four different cancer indications, specifically non-small cell lung cancer, breast cancer, colorectal cancer, and multiple myeloma. It will place a special emphasis on systematic evaluation of fitness of data in relation to the study design and statistical analysis for a particular research question, and pre-registration of study protocols prior to initiation and analysis. Agreement of treatment effect estimates between RCTs and their respective emulations will be assessed using pre-specified criteria. Through extensive sensitivity analyses benchmarked against RCT results, the ENCORE project will inform understanding of how measurement, design, and analytic decisions influence the interpretation of results from emulated oncology trials using RWD.

@@

Randomized controlled trials (RCTs) are the evidentiary gold standard methodology for establishing the efficacy and safety of medical products. Guided by the 21^st^ Century Cures Act,[@RWEFDA] the Food and Drug Administration (FDA) established a framework to factor considerations for use of real-world evidence (RWE) generated from real-world data (RWD) such as electronic health records (EHR) to support approval of new indications or satisfy postmarketing requirements.[@RWEFDA] Accounting for approximately 30% of new FDA drug approvals, oncology was the disease area with the most new approvals in 2024[@mullard20242023] as well as several indication expansions, and has many areas of high unmet medical need to treat serious conditions. Therefore, RWE can have a particularly important potential to complement evidence from RCTs in the field of oncology. Potential uses include the assessment of effectiveness in specific patient populations that are not adequately represented in RCTs, or the precision oncology-focused discovery of biomarkers among pan-tumor populations that harbor specific genomic and immuno-pathological signatures.

However, the comparability and transportability of results derived between observational studies and RCTs depends on multiple factors. Frequently referenced limitations include lack of baseline randomization and the imbalance in prognostic factors resulting thereof, missing data, unmeasured confounding, small study sizes, data discontinuity,[@Merola2022; @joshua2022longitudinal] adoption of changes in treatment guidelines over time in routine clinical care and the inability to measure and emulate common eligibility criteria, including prognostic factors, and standardized response assessments in real-world data (RWD).[@rider2024emulations] While examples of oncology trial emulations have been published,[@rider2024emulations; @merola2023aetion; @merola2024calibrating] a systematic and scaled approach to emulate a diverse set of oncology trials with multiple heterogeneous databases is necessary to increase confidence in the interpretability of observational studies, evaluate regulatory considerations, and to provide context as to which questions can be validly answered.

The RCT DUPLICATE initiative[@wang2023emulation] increased our understanding of when observational studies can come to causal conclusions on treatment effects by benchmarking results against RCTs under the assumption that each RCT finding reflects a causal treatment effect. In settings where the RCT designs could be emulated well, observational studies were able to reach similar conclusions.[@heyard2024design] However, this prior work from RCT-DUPLICATE focused primarily on emulating trials in non-oncology settings using claims databases.

The *Emulation of Comparative Oncology Trials with Real-world Evidence* (ENCORE) project [@encoreFDA] aims to extend this work to oncology. Clinical studies in oncology come with unique methodological challenges which must be systematically explored and understood. Building on a process co-developed with the FDA through RCT DUPLICATE,[@wang2023emulation] this expansion to oncology will emulate 12 randomized clinical trials using multiple specialty oncology EHR data sources. The process will emphasize transparency and include a data fitness-for-use assessment[@rivera2024oncology; @gatto2022structured] for each RWD source with respect to each trial emulation as well as extensive sensitivity analyses to assess robustness of findings.

The objectives of this project are to 1) develop state-of-the-art methodological approaches and 2) apply these methods to gain insights into the potential use of RWE to enhance regulatory and clinical decision-making in oncology. To achieve these objectives, this demonstration project will systematically emulate 12 oncology trials across four cancer types and assess the agreement of treatment effect estimates between RCTs and their respective emulations.

Here, we describe the design and process for the selection of the 12 oncology RCTs, assessment of the database quality and selection, protocol development and pre-registration, study design and statistical analysis plans, and pre-specified agreement metrics to evaluate the concordance between RCTs and emulations.

@@

The focus of ENCORE is to efficiently explore in what contexts observational studies can or cannot yield similar results compared to oncology RCTs. Therefore, the project emphasizes trials of therapies for common cancers for which there has been substantial therapeutic development in recent years. After review and collaboration with clinical and regulatory experts, four cancer indications were identified: non-small cell lung cancer, breast cancer, colorectal cancer and multiple myeloma. For each cancer type, we aim to conduct three trial emulations using multiple accessible databases (i.e., the total equaling 12 trial emulations x *n* databases which are found fit-for-purpose for each trial).

We used a systematic process for trial selection where the eligibility criteria are documented in CONSORT diagrams showing reasons for excluding RCTs by cancer type. The search was conducted using the *Aggregated Analysis of ClinicalTrials.gov* (AACT) database which is a publicly available relational database developed and maintained by the Clinical Trials Transformation Initiative (CTTI) that contains information (protocol and result data elements) about studies registered on ClinicalTrials.gov.[@tasneem2012database] To identify eligible trials, we used a combined search query strategy of the National Library of Medicine (NLM)-controlled *Medical Subject Headings **(***MeSH) term and a free keyword search for the respective cancer indication in the *conditions*, *studies* and *detailed_descriptions* fields of each trial entry on ClinicalTrials.gov.

@@

-   Intervention model: Parallel assignment

@@

-   Trial start: 2011 or later

-   Primary purpose: Study treatment efficacy

-   Endpoint(s): Overall survival must be one of the endpoints reported (either as hazard ratio or median overall survival time)

@@

The rationale and operationalization for each criterion is listed in detail in @tbl-criteria. We mainly considered pivotal large late-phase RCTs that were initiated after 2011 because treatment guidelines for included cancer indications have undergone significant changes in recent years. Due to the rapid adoption of newly-approved therapies in routine care, patients are less likely to receive outdated treatment regimens in clinical practice. Conversely, we excluded trials with results published too recently to allow for substantive medical product uptake and follow-up time accrual in real-world databases used for this project. We did not define a global cut-off as the requirements for follow-up time are different for each cancer type and population (e.g., advanced NSCLC versus early-stage breast cancer) and the decisions were made on a clinically relevant basis.

Although there have been substantial methodological advancements to increase our understanding on the emulation and comparison of real-world progression-free survival (PFS) and objective response rates (ORR) to a RECISTv1.1[@eisenhauer2009new]-based PFS and ORR assessment in RCTs [@ThanCCR; @mckelvey2024evaluation], imaging-based evaluations still require a level of granularity which is not well reflected in chart-abstracted assessments of a patient's progression in routine care.[@rwdRECIST; @rivera2022friends] The timing and cadence of intervals between progression assessments can differ between RCTs and routine care which may result in measurement error and bias.[@ackerman2024measurement] Given this, and the large number of other methodological challenges including missing data, small sample sizes, data discontinuity and rapidly changing guideline treatments, we focus on the emulation of overall survival (OS) as the endpoint of interest. Therefore, we include only trials that have reported OS as a pre-specified endpoint in the protocol.

While most trial eligibility criteria could be operationalized in an automated fashion, our final criterion was an assessment of emulation feasibility and clinical relevance. This criterion involved extensive human review. The fundamental points considered in this step include an initial feasibility assessment of the data fitness,[@rivera2022friends] involving an assessment of whether critical eligibility criteria (e.g., biomarker status) including prognostic factors (e.g., Eastern Cooperative Oncology Group \[ECOG\] performance score) are measurable, and whether preliminary study size based on a rough estimation of the number of patients observed in the data with the combination of treatment regimen and line of therapy is reasonable for shortlisting the trial. Lastly, trial candidates were ranked and shortlisted as primary or runner-up based on their clinical and regulatory relevance.

A list of tentative, shortlisted primary candidates is presented in @tbl-rcts. The corresponding selection process is illustrated in the CONSORT diagrams (**Supplementary Figures 1-4**) and runner up candidate trials are listed in **Supplementary Table 1**. The majority of shortlisted trials represent indications for advanced (locally advanced, inoperable, recurrent/progressive disease) or metastatic cancer patients because a large proportion of drug development efforts have recently focused on these settings. A key objective that we aim to explore with the selected trials is to evaluate how stage (early, late), line of therapy (\[neo\]adjuvant, first line, advanced lines of therapy), therapy protocols (monotherapy, combination therapy) and clinical characteristics (all-comer versus complex genetic or immunological signatures) can be emulated using RWD. If more thorough feasibility assessments after applying all relevant eligibility criteria suggest that the threat of bias from mis-measurement of key study parameters or residual confounding remains high, or that the study size is not sufficient, then runner-up candidate trials will be considered instead.

@@

The ENCORE project will use data from four U.S.-based oncology-specific EHR-derived RWD data sources (in alphabetical order): ConcertAI, COTA, Flatiron Health, and Ontada/McKesson. A detailed description and sampling methodology will be provided with each trial emulation protocol. For ENCORE, not all databases will be available for each cancer indication given the specificity of data elements required and the names of the databases will be blinded and referred to as ENCORE DataBase (EDB) 1, 2, 3 and 4 for the final reporting of results (in randomly assigned order). If more than one database is considered fit-for-use for a respective trial emulation, the most suitable analytic model will be employed for each database separately and final treatment effect estimates will be pooled using a meta-analytic approach using fixed effects (weights reflect the inverse of the variance of each database study) and random effects (weights reflect the inverse of the variance of each database study plus an additional variance term that quantifies the assumed heterogeneity between databases) models.[@nikolakopoulou2014interpret]

@@

For each selected RCT, a detailed protocol, pre-specifying key elements of the trial emulation, will be developed using the HARPER protocol template[@wang2022harmonized;@guidelinegeneral] and will be registered on ClinicalTrials.gov after review by an expert panel comprising clinicians and FDA reveiwers. Following the target trial emulation framework, we will provide an explicit description and rationale on how each element will be emulated including database selection, covariate measurement, operationalization of key eligibility criteria, study design, data analysis and causal contrasts of interest.[@hernan2022target; @hernan2016specifying] Since it is common that oncology RCTs update survival estimates periodically based on accrued follow-up time, the protocol will give a brief summary of each emulated RCT and specify which target OS estimates will be used for agreement metric evaluation (see @sec-agreement-metrics). All eligibility criteria will be extracted based on publications, publicly available protocols, and statistical analysis plans of the selected RCT.

@@

An important aspect when emulating oncology trials is the choice and estimation of the appropriate estimand of interest.[@rufibach2018] In prior RCT-DUPLICATE efforts, a “while on treatment” strategy was chosen for database studies that were designed to emulate trials where the primary trial analysis was intention-to-treat (e.g., a “treatment policy” estimand). This estimand was chosen to mimic the high adherence typically observed in large cardiovascular clinical trials in the context of low adherence in clinical practice. However, a “while on treatment” estimand would be inappropriate when emulating oncology trials, not only because of highly informative censoring, but also because OS outcome analyses in oncology are typically “treatment policy” estimands that allow for crossover or other post-progression antineoplastic therapy.[@manitz2022estimands] We plan to focus on “treatment policy” estimands for the primary analyses in our trial emulations. However, allowing crossover can dilute treatment effects, moving point estimates toward the null. Therefore, differences in cross-over and discontinuation rates between the trial and emulation may still contribute to differences in observed outcomes. Recognizing these practical challenges, we will carefully characterize the patterns of discontinuation or cross-over in both the trials and the database studies designed to emulate them.

@@

Real-world data fitness and emulation feasibility for each candidate trial will be assessed in multiple steps based on the approach described by the Oncology Quality, Characterization, and Assessment of Real-World Data (Oncology QCARD) Initiative.[@rivera2022friends] The first step assesses if relevant variables like exposure, line of therapy, outcomes, and covariates are generally available, measured and operationalizable in routine-care. Since many oncology RCTs in recent years have focused on biomarker-defined populations, nuances in measurement and operationalizability of specific biomarkers must be understood to ensure a representative and adequately sized study population. For example, immune checkpoint inhibitors have significantly changed the cancer treatment landscape since the first approvals of the CTLA-4 inhibitor ipilimumab in 2011 and the PD-1 inhibitors pembrolizumab and nivolumab in 2014 in the United States. However, the operationalization of the expression of the PD-L1 biomarker in RCTs (e.g., percent staining, tumor proportion score, or combined positive score) has evolved, and as such PD-L1 '*positivity'* may have different definitions by year based on different cut-off values.

According to the *Structured Process to Identify Fit-For-Purpose Data* (SPIFD) framework[@gatto2022structured], the next step outlines how eligibility criteria will be ascertained using a color-coded heatmap that will indicate the level of confidence on how well each criterion can be emulated in each selected database. As there are eligibility criteria in oncology clinical trials which either are infeasible to emulate (e.g., physician-assessed survival prognosis quantification) or that do not impact clinical care for the emulation of the trial (e.g., male patients should be willing to use barrier contraception), the study team will determine key eligibility criteria for the emulation of the trial based on consensus.

Additionally, we will provide conceptual and operationalized definitions on how exposures, outcomes and covariates are defined in each respective database. There will be special emphasis on how exposure, in context of disease and line of therapy settings, and the survival outcomes are emulated. For all considered databases, real-world OS (rwOS) is typically a composite measure that, depending on the underlying database, can be derived from different mortality sources (i.e., EHR documentation, Social Security Death Index, and other linkages). Given that not all relevant sources that provide mortality data are synchronized and updated uniformly, sensitivity analyses with more conservative (i.e., earlier) censoring dates will be considered for each trial emulation to mitigate the potential impact of ghost-time bias.[@meyer2020open]

@@

Particularly when emulating pivotal trials of practice-changing treatments, multiple aspects need to be considered, such as the contemporaneity of the control cohort, the adoption rate of the novel medical product in routine care, the magnitude of the clinical treatment benefit, and the rate in which patients discontinue or cross-over, as they could influence effect estimates.

To that end, comprehensive data explorations will be performed as part of the protocol development to contextualize these parameters and (if reported) evaluate comparisons to the emulated trial. Examples for such standard diagnostics are visualized in @fig-initiators.

The distribution of patient characteristics, by exposure status, will be examined before and after applying eligibility criteria and contrasted with the distributions of patient characteristics of the original RCT. Initial propensity score matching or weighting methods will be applied to ensure that measured pre-exposure covariates can be balanced, exposure cohorts are conditionally exchangeable at baseline, and resulting sample sizes are still sufficient after matching or weighting. At this stage, all exploratory analyses will be conducted blinded towards the outcome to not bias any study design or analytic choices based on known outcome information.

@@

Causal analyses of observational data are often not designed with respect to formal hypothesis testing and statistical power in the same manner as RCTs since the number of 'recruited' patients is limited by available data .[@hernan2022causal] For this project, however, the emulations in EHR data must have at least equal power to the relevant trial for interpretation of the pre-specified agreement metrics between RCT and RWD results. Since the main outcome of interest is defined as time from index treatment initiation to all-cause mortality (rwOS), the estimation of the statistical power is driven by the number of events rather than the number of patients. To assess if the overall number of events, unstratified by exposure, is sufficient such that a significant difference can be detected based on the original RCT-reported hazard ratio (HR), the statistical power (1-$\beta$) will be estimated using Schoenfeld's sample-size formula for the proportional-hazards regression model.[@schoenfeld1981asymptotic]

@@

To formally compare treatment effects between RCTs and their respective emulations, we will adapt the approach of the RCT-DUPLICATE project.[@franklin2020nonrandomized; @wang2023emulation] That is, for the primary endpoint of interest (hazard ratio \[HR\] for OS and corresponding 95% confidence intervals), we will derive three qualitative agreement metrics: statistical agreement, estimate agreement and agreement based on the standardized mean difference (SMD). Examples are illustrated in @tbl-metrics.

**Statistical significance agreement**: agreement between RCT and emulated trial treatment effect with regards to directionality and statistical significance (nominal in the case of emulated trials).

**Estimate agreement**: agreement that the estimated RWE treatment effect is within the 95% CI of the RCT treatment effect estimate. Provided that for some emulations, the power of the RWE study may be larger than that of the original RCT, this could lead to situations where there is no statistical significance agreement (RCT estimate crossing the null although the treatment effect estimates are overlapping) or vice versa in case the RCT has a larger power than the RWE emulation.

@@

The study design for each trial emulation will be visualized as part of the protocol using a graphical depiction of the exact measurement windows of eligibility criteria, washout periods, and covariates relative to the cohort entry time.[@schneeweiss2019graphical]

@@

Due to its frequency of use in oncology trials, the primary parameter of interest in ENCORE will be defined as the marginal HR coefficient for the treatment comparison for OS.[@cox1972regression] We will also consider alternative endpoints on an absolute risk scale such as median survival times, survival probabilities at pre-defined time points during follow-up[@kaplan1958nonparametric] and restricted mean survival times as secondary endpoints of interest.
